---
title: "Mick's project file"
output: html_document
date: "2026-03-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

Does temperature affect collision numbers? 
Visibility and collisions (barplot??? boxplot???) … acknowledge not having bunch of “high” vs. “low” visibility → could be a separate table… how much data set is in each. 

```{r, echo=FALSE}
library("tidyverse")
library("janitor")
library(RColorBrewer)
library(paletteer)
library(ggthemes)
```

```{r, echo=FALSE}
birds <- read.csv("data/bird_collision.csv")
```

##colors
```{r}
colors <- paletteer::palettes_d_names
```
```{r}
my_palette <- paletteer_d("ggsci::indigo_material")
my_palette
```


##Separate date column

```{r}
birds <- birds %>% separate(col="date", into = c("month","day","year"), sep = "/") %>% ##separate month into col
  select(!day) %>% ##remove day, since data already has a day col
  mutate(month=as.factor(month)) ##make month into a factor
```

##Add season variable

```{r}
birds <- birds %>% ##make a new column showing what season, since the data is split by spring or fall
  mutate(season=case_when(
    month=="4" | month=="5" | month=="6" ~ "spring",
    month=="8" | month=="9" | month=="10" | month=="11" ~ "fall",
    .default=NA
  ))
```

##Adding migration level variable based on the paper

```{r}
birds <- birds %>% 
  mutate(migration_level= case_when( ##migration level from the paper
    season=="spring" & amt>=4 & amt<=1860 ~ "medium", ##946 is average migration for spring
    season=="fall" & amt>=9 & amt<=3420 ~ "medium", ##1690 is avg for fall
    season=="spring" & amt<=4 ~ "low", ##avg-1SD
    season=="fall" & amt<=9 ~ "low",
  season=="spring" & amt>=1860 ~ "high", ##avg+1SD
    season=="fall" & amt>=3420 ~ "high"
  ))
```

##Separate wind speed into low, med, high with same method as migration level: low<= avg-1SD, med<= avg+1SD, >= avg-1SD, high>=avg+1SD. I took the absolute value of the winds first, since negative wind is just the opposite direction of positive, and I want to compare strength, not direction

```{r}
birds %>% 
  group_by(season) %>% 
  mutate(uwind_abs=abs(uwind), vwind_abs=abs(vwind)) %>% 
  summarise(mean(uwind_abs), sd(uwind_abs), mean(vwind_abs), sd(vwind_abs))
```

```{r}
birds <- birds %>% 
  mutate(uwind_abs=abs(uwind), vwind_abs=abs(vwind)) %>%
  mutate(uwind_level= case_when(
    season=="spring" & uwind_abs<= 2.457172-2.132746 ~"low",
    season=="fall" & uwind_abs<= 2.184764-1.760663 ~"low",
    season=="spring" & uwind_abs>= 2.457172+2.132746 ~"high",
    season=="fall" & uwind_abs>= 2.184764+1.760663 ~"high",
    season=="spring" & uwind_abs>= 2.457172-2.132746 & uwind<= 2.457172+2.132746 ~"medium",
    season=="fall" & uwind_abs>= 2.184764-1.760663 & uwind_abs<= 2.184764+1.760663 ~"medium",
  )) %>% 
  mutate(vwind_level= case_when(
    season=="spring" & vwind_abs<= 2.648716-1.587146 ~"low",
    season=="fall" & vwind_abs<= 2.357720-1.516771 ~"low",
    season=="spring" & vwind_abs>= 2.648716+1.587146 ~"high",
    season=="fall" & vwind_abs>= 2.357720+1.516771 ~"high",
    season=="spring" & vwind_abs>= 2.648716-1.587146 & vwind<= 2.648716+1.587146 ~"medium",
    season=="fall" & vwind_abs>= 2.357720-1.516771 & vwind_abs<= 2.357720+1.516771 ~"medium",
  ))
```

```{r}
names(birds)
```

##Does wind affect collision numbers? (do separate for uwind and vwind) 

##uwind spring

```{r, echo=FALSE}
birds %>% 
  filter(season=="spring") %>% 
  ggplot(aes(x=uwind, y=count, color=migration_level))+
  geom_point()+
  geom_smooth(method=lm, color="black", se=T)+
  facet_wrap(~migration_level)+
  theme(legend.position = "bottom")+
  labs(y="Number of collisions",x="East-west wind speed (m/s)", title="East-west wind speed compared to collision count in spring, separated by migration level", color="Migration level")+
  theme_tufte()+
  scale_fill_manual(values=my_palette[3:10])+ ##first two colors were too light for a scatterplot, so i removed them here
  scale_color_manual(values=my_palette[3:10])
```

##uwind fall

```{r, echo=FALSE}
birds %>% 
  filter(season=="fall") %>% 
  ggplot(aes(x=uwind, y=count, color=migration_level))+
  geom_point()+
  geom_smooth(method=lm, color="black", se=T)+
  facet_wrap(~migration_level)+
  scale_y_log10()+
  labs(y="Number of collisions",x="East-west wind speed (m/s)", title="East-west wind speed compared to collision count in fall, separated by migration level", color="Migration level")+
  theme_tufte()+
  scale_fill_manual(values=my_palette[3:10])+ ##first two colors were too light for a scatterplot, so i removed them here
  scale_color_manual(values=my_palette[3:10])
```

##vwind spring

```{r, echo=FALSE}
birds %>% 
  filter(season=="spring") %>% 
  ggplot(aes(x=vwind, y=count, color=migration_level))+
  geom_point()+
  geom_smooth(method=lm, color="black", se=T)+
  facet_wrap(~migration_level)+
  theme(legend.position = "bottom")+
  labs(y="Number of collisions",x="North-south wind speed (m/s)", title="North-south wind speed compared to collision count in spring, separated by migration level", color="Migration level")+
  theme_tufte()+
  scale_fill_manual(values=my_palette[3:10])+ ##first two colors were too light for a scatterplot, so i removed them here
  scale_color_manual(values=my_palette[3:10])
```

##vwind fall

```{r, echo=FALSE}
birds %>% 
  filter(season=="fall") %>% 
  ggplot(aes(x=vwind, y=count, color=migration_level))+
  geom_point()+
  geom_smooth(method=lm, color="black", se=T)+
  facet_wrap(~migration_level)+
  scale_y_log10()+
  labs(y="Number of collisions",x="North-south wind speed (m/s)", title="North-south wind speed compared to collision count in fall, separated by migration level", color="Migration level")+
  theme_tufte()+
  scale_fill_manual(values=my_palette[3:10])+ ##first two colors were too light for a scatterplot, so i removed them here
  scale_color_manual(values=my_palette[3:10])
```

##vwind spring

```{r, echo=FALSE}
birds %>% 
  filter(season=="spring") %>% 
  ggplot(aes(x=vwind, y=count, color=migration_level))+
  geom_point()+
  geom_smooth(method=lm, color="black", se=T)+
  facet_wrap(~migration_level)+
  scale_y_log10()+
  labs(y="Number of collisions",x="North-south wind speed (m/s)", title="North-south wind speed compared to collision count in spring, separated by migration level", color="Migration level")+
  theme_tufte()+
  scale_fill_manual(values=my_palette[3:10])+ ##first two colors were too light for a scatterplot, so i removed them here
  scale_color_manual(values=my_palette[3:10])
```


##boxplots

##uwind fall

```{r, echo=FALSE}
birds %>% 
  filter(season=="fall") %>% 
  ggplot(aes(x=uwind_level, y=count, fill=migration_level, position="dodge"))+
  geom_boxplot()+
  #facet_wrap(~migration_level)+
  scale_y_log10()+
  labs(y="Number of collisions",x="East-west wind level", title="East-west wind speed compared to collision count in fall, separated by migration level", fill="Migration level")+
  theme_tufte()+
  scale_fill_manual(values=my_palette)+
  scale_color_manual(values=my_palette)
```

##uwind spring

```{r, echo=FALSE}
birds %>% 
  filter(season=="spring") %>% 
  ggplot(aes(x=uwind_level, y=count, fill=migration_level, position="dodge"))+
  geom_boxplot()+
  #facet_wrap(~migration_level)+
  scale_y_log10()+
  labs(y="Number of collisions",x="East-west wind level", title="East-west wind speed compared to collision count in spring, separated by migration level", fill="Migration level")+
  theme_tufte()+
  scale_fill_manual(values=my_palette)+
  scale_color_manual(values=my_palette)
```

##vwind fall

```{r, echo=FALSE}
birds %>% 
  filter(season=="fall") %>% 
  ggplot(aes(x=vwind_level, y=count, fill=migration_level, position="dodge"))+
  geom_boxplot()+
  #facet_wrap(~migration_level)+
  scale_y_log10()+
  labs(y="Number of collisions",x="North-south wind level", title="North-south wind speed compared to collision count in fall, separated by migration level", fill="Migration level")+
  theme_tufte()+
  scale_fill_manual(values=my_palette)+
  scale_color_manual(values=my_palette)
```

##vwind spring

```{r, echo=FALSE}
birds %>% 
  filter(season=="spring") %>% 
  ggplot(aes(x=vwind_level, y=count, fill=migration_level, position="dodge"))+
  geom_boxplot()+
  #facet_wrap(~migration_level)+
  scale_y_log10()+
  labs(y="Number of collisions",x="North-south wind level", title="North-south wind speed compared to collision count in spring, separated by migration level", fill="Migration level")+
  theme_tufte()+
  scale_fill_manual(values=my_palette)+
  scale_color_manual(values=my_palette)
```

